\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-05-31 submission/manuscript.Rmd
update manuscript after Jennas edits?

---
title: "**Evaluation of machine learning methods for 16S rRNA gene data**"
bibliography: references.bib
output:
  pdf_document: 
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

@@

Machine learning (ML) modeling of the human microbiome is used to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but reliable ML models are scarce in part to the flawed modeling methods. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained the fastest (12 min). This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

## Introduction

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is difficult as human associated microbial communities are remarkably complex, high-dimensional and uneven within and between individuals with the same disease. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for differences in health outcomes. Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; dadkhah_gut_2019]. More notably, flawed ML practices are prevalent such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance, reporting few or only the best outcomes of different randomizations of cross-validation and showing a disregard for large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017]. Moreover, there is a lack of discussion on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018]. The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability. We need to strive toward better machine learning practices by (1) implementing consistent and reliable machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

To showcase a reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients. We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. Our ML pipeline utilized held-out test data to evaluate generalization performance of each ML model. The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`. Random forest had the highest median AUROC for detecting SRN. Despite the simplicity, the L2-regularized logistic regression followed random forest in performance. In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[7],3], 3))`). We found that median cross-validation and testing AUROC varied only `r paste(round(abs(difference[difference_index[7]]), 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

__Model performance and generalizability.__

We evaluated the prediction performances of seven binary classification models when applied to held-out test data using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric [Figure 1]. Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p = `r paste(round(rf_logit$p.value, 4))`). The median AUROC of random forest was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2 logistic regression, XGBoost, L2 SVM with linear and radial basis function kernel AUROC values were not significantly different from one another. They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively. L1 SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].

@@

#### Interpretation of results
 - In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome.

 - We showed the importance of held-out test set and reporting the results of many iterations of cross-validation and testing. Each data-split has a weight that comes from random splitting. Seeing that testing and cross-validation median differences being low show that the models are generalizable and the pipeline works.

 - Our results also show that we can choose different models for different reasons. Getting feature importances as weights and regression coefficients is the fastest and easiest interpratation. Similarly, linear models require less computational burden, fastest to train. Thinking about the applicability, and actionability of a model in clinical setting, we might want faster results.

 -This is a huge difference which shows that you can get "lucky" or "unlucky" depending on the random split. It is crucial to report all the findings, many randomizations of these splits to get reliable performance metrics.

>>>>

---
title: '**Evaluation of machine learning methods for 16S rRNA gene data**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

@@

Machine learning (ML) modeling of the human microbiome is used to identify the microbial biomarkers and aid in diagnosis of many chronic diseases such as inflammatory bowel disease, diabetes and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes from bacterial abundances, but reliable ML models are scarce in part due to the flawed modeling methods. Furthermore, using black box ML models has hindered the validation of microbial biomarkers. To overcome these challenges, we benchmarked seven different ML models that use fecal 16S rRNA sequences to predict the presence/absence of colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229 cases). To show the effect of model selection, we assessed the accuracy, interpretability, and computational efficiency of the following models: L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model was best at detecting CRC lesions with an AUROC of 0.695 but it was slow to train (101.3 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, and it trained the fastest (12 min). This study showed that we should choose ML models based on our expectations of performance and interpretability as well as our  computational resources. It also established standards for modeling pipelines of microbiome-associated ML models.

@@

## Background

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences have allowed rapid exploration of human associated microbiome and its health implications. Currently, the human microbiome field is growing at an unprecedented rate and as a result, there is an increasing demand for methods that identify associations between members of the microbiome and human health. However, this is difficult as human associated microbial communities are remarkably complex, high-dimensional and uneven within and between individuals with the same disease. It is unlikely that a single species can explain a disease. Instead, subsets of those communities, in relation to one another and to their host, account for differences in health outcomes.

Machine learning (ML) methods are effective at recognizing and highlighting patterns in complex microbial datasets. Therefore, researchers have started to explore the utility of ML models that use microbiota associated biomarkers to predict human health and to understand the microbial ecology of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018].  However, currently the field's use of ML lacks clarity and consistency on which methods are used and how these methods are implemented [@thaiss_persistent_2016; dadkhah_gut_2019]. More notably, flawed ML practices are prevalent such as using ML pipelines where there is no seperate held-out test dataset to evaluate model performance, reporting few or only the best outcomes of different randomizations of cross-validation and showing a disregard for large differences between cross-validation and testing performances as well as large confidence intervals of testing performances [@flemer_oral_2018; @baxter_microbiota-based_2016; @dai_multi-cohort_2018; @montassier_pretreatment_2016; @papa_non-invasive_2012; @mossotto_classification_2017; @ai_systematic_2017; @wong_quantitation_2017]. Moreover, there is a lack of discussion on why a particular ML model is utilized. Recently, there is a trend towards using more complex ML models such as random forest, extreme gradient boosting and neural networks without a discussion on if and how much model interpretibility is necessary for the study [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017; @ @geman_deep_2018].

The lack of transparency on modeling methodology and model selection negatively impact model reproducibility and reliability. We need to strive toward better machine learning practices by (1) implementing consistent and reliable machine learning pipelines and (2) selecting ML models that reflect the goal of the study as it will inform our expectations of model accuracy, complexity, interpretibility and computational efficiency.

To showcase a reliable ML pipeline and to shed light on how much ML model selection can affect modeling results, we performed an empirical analysis comparing several different ML models using the same dataset and with a robust ML pipeline. We used a previously published colorectal cancer (CRC) study [@baxter_dna_2016] which had fecal 16S rRNA gene sequences from `r paste(nrow(shared))` patients. CRC is the third most common form of cancer worldwide in which the human-associated human microbiome is hypothesized to directly contribute to its development. We built ML models using fecal 16S rRNA gene sequences to predict patients with normal colons or patients with colonic tumors which are called screen relevant neoplasias (SRN). The study had `r paste(sum(sum(meta$dx=="normal")+sum(meta$Dx_Bin=="Adenoma")))` normal and `r paste(sum(sum(meta$dx=="cancer")+sum(meta$Dx_Bin=="adv Adenoma")))` SRN samples. We established modeling pipelines for L2-regularized logistic regression, L1 and L2 support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest and XGBoost. Our ML pipeline utilized held-out test data to evaluate generalization performance of each ML model. The median test AUROC varied from `r paste(round(test_all[performance_index[1],3], 3))`  to `r paste(round(test_all[performance_index[7],3], 3))`. Random forest had the highest median AUROC for detecting SRN. Despite its simplicity, the L2-regularized logistic regression followed random forest in performance. In terms of computational efficiency, L2 logistic regression trained the fastest (`r paste(round(walltime_df[walltime_index[1],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[1],3], 3))`), while XGBoost took the longest (`r paste(round(walltime_df[walltime_index[7],2], 3))` hours, std ± `r paste(round(walltime_df[walltime_index[7],3], 3))`). We found that median cross-validation and testing AUROC varied `r paste(round(abs(difference[difference_index[7]]), 3))`, which highlights the importance of a seperate held-out test set and consistent preprocessing of the data prior to evaluation. Aside from evaluating generalization and classification performances for each of these models, this study established standards for modeling pipelines of microbiome-associated machine learning models.

@@

__Discriminative performance of the seven models.__

We evaluated the prediction performances of seven binary classification models when applied to held-out test data using the area under the receiver operating characteristic curve (AUROC) as the discriminative performance metric [Figure 1]. Random forest had significantly higher test AUROC values than the other models for detecting SRNs when AUROC values were compared to the other six by Wilcoxon rank sum test (p = `r paste(round(rf_logit$p.value, 4))`). The median AUROC of therandom forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another. They had median AUROC values of `r paste(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`) respectively. L1 SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2].

__Generalizability of the seven models.__

@@

#### Interpretation of results

- In this study we established a robust ML pipeline to use 16S rRNA sequence counts to predict a binary health outcome. We used a held-out test set to illustrate the difference between cross-validation and testing AUROC values. When the differences between cross-validation and test performance is low, this suggets the models will perform similar with similar data. In all seven models, the differences in cross-validation and testing AUROC values did not exceed `r paste(round(abs(difference_model[difference_index[7],1]), 3))` which suggests that these models will be able to test similar new data.

- We also reported the results of 100 random dataset splits because there are differences in AUROC values of each of the seven models between different random data-splits. Our results showed that the testing AUROC values varied `r paste(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 2))` on average, depending on the data-split. This result shows that you can get "lucky" in terms of prediction performance depending on the random split. Therefore, it is crucial to report all the findings over many random data-splits to get reliable performance metrics.

 - Our results also showed that we can choose different models for different reasons. Getting feature importances as weights and regression coefficients is the fastest and easiest interpratation. Similarly, linear models require less computational burden, fastest to train. Thinking about the applicability, and actionability of a model in clinical setting, we might want faster results.

 -

### 2019-05-31 submission/manuscript.tex
update manuscript after Jennas edits?

Machine learning (ML) modeling of the human microbiome is used to
identify the microbial biomarkers and aid in diagnosis of many chronic
diseases such as inflammatory bowel disease, diabetes and colorectal
cancer. Progress has been made towards developing ML models that predict
health outcomes from bacterial abundances, but reliable ML models are
scarce in part to the flawed modeling methods. Furthermore, using black
box ML models has hindered the validation of microbial biomarkers. To
overcome these challenges, we benchmarked seven different ML models that
use fecal 16S rRNA sequences to predict the presence/absence of
colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229
cases). To show the effect of model selection, we assessed the accuracy,
interpretability, and computational efficiency of the following models:
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and extreme gradient boosting (XGBoost). The random forest
model was best at detecting CRC lesions with an AUROC of 0.695 but it
was slow to train (101.3 h) and hard to interpret. Despite its
simplicity, L2-regularized logistic regression followed random forest in
predictive performance with an AUROC of 0.680, and it trained the
fastest (12 min). This study showed that we should choose ML models
based on our expectations of accuracy, interpretability and
computational efficiency. It also established standards for modeling
pipelines of microbiome-associated ML models.

@@

\subsection{Introduction}\label{introduction}

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is difficult as human
associated microbial communities are remarkably complex,
high-dimensional and uneven within and between individuals with the same
disease. It is unlikely that a single species can explain a disease.
Instead, subsets of those communities, in relation to one another and to
their host, account for differences in health outcomes. Machine learning
(ML) methods are effective at recognizing and highlighting patterns in
complex microbial datasets. Therefore, researchers have started to
explore the utility of ML models that use microbiota associated
biomarkers to predict human health and to understand the microbial
ecology of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
(1--11). However, currently the field's use of ML lacks clarity and
consistency on which methods are used and how these methods are
implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, flawed ML
practices are prevalent such as using ML pipelines where there is no
seperate held-out test dataset to evaluate model performance, reporting
few or only the best outcomes of different randomizations of
cross-validation and showing a disregard for large differences between
cross-validation and testing performances as well as large confidence
intervals of testing performances (4, 13--19). Moreover, there is a lack
of discussion on why a particular ML model is utilized. Recently, there
is a trend towards using more complex ML models such as random forest,
extreme gradient boosting and neural networks without a discussion on if
and how much model interpretibility is necessary for the study (11,
20--22). The lack of transparency on modeling methodology and model
selection negatively impact model reproducibility and reliability. We
need to strive toward better machine learning practices by (1)
implementing consistent and reliable machine learning pipelines and (2)
selecting ML models that reflect the goal of the study as it will inform
our expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase a reliable ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and with a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. We built ML models using fecal 16S rRNA gene sequences to
predict patients with normal colons or patients with colonic tumors
which are called screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate generalization performance of each ML model. The median test
AUROC varied from 0.601 to 0.695. Random forest had the highest median
AUROC for detecting SRN. Despite the simplicity, the L2-regularized
logistic regression followed random forest in performance. In terms of
computational efficiency, L2 logistic regression trained the fastest
(0.202 hours, std ± 0.019), while XGBoost took the longest (162.843
hours, std ± 3.986). We found that median cross-validation and testing
AUROC varied only 0.021, which highlights the importance of a seperate
held-out test set and consistent preprocessing of the data prior to
evaluation. Aside from evaluating generalization and classification
performances for each of these models, this study established standards
for modeling pipelines of microbiome-associated machine learning models.

@@

\textbf{Model performance and generalizability.}

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using the area under the
receiver operating characteristic curve (AUROC) as the discriminative
performance metric {[}Figure 1{]}. Random forest had significantly
higher test AUROC values than the other models for detecting SRNs when
AUROC values were compared to the other six by Wilcoxon rank sum test (p
= 0.0011). The median AUROC of random forest was 0.695 (IQR 0.044). L2
logistic regression, XGBoost, L2 SVM with linear and radial basis
function kernel AUROC values were not significantly different from one
another. They had median AUROC values of 0.68 (IQR 0.055), 0.679 (IQR
0.053), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1 SVM
with linear kernel and decision tree had significantly lower AUROC
values than the other ML models with median AUROC of 0.65 (IQR 0.066)
and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.

@@

\begin{itemize}
\item
  In this study we established a robust ML pipeline to use 16S rRNA
  sequence counts to predict a binary health outcome.
\item
  We showed the importance of held-out test set and reporting the
  results of many iterations of cross-validation and testing. Each
  data-split has a weight that comes from random splitting. Seeing that
  testing and cross-validation median differences being low show that
  the models are generalizable and the pipeline works.
\item
  Our results also show that we can choose different models for
  different reasons. Getting feature importances as weights and
  regression coefficients is the fastest and easiest interpratation.
  Similarly, linear models require less computational burden, fastest to
  train. Thinking about the applicability, and actionability of a model
  in clinical setting, we might want faster results.
\end{itemize}

-This is a huge difference which shows that you can get ``lucky'' or
``unlucky'' depending on the random split. It is crucial to report all
the findings, many randomizations of these splits to get reliable
performance metrics.

\begin{itemize}
\item
  Suprisingly, l2 logistic regression does really well.
\item
  Computational efficiency and interp of logistic vs random forest
\item
  decision tree and l1 why so low?
\end{itemize}

>>>>

Machine learning (ML) modeling of the human microbiome is used to
identify the microbial biomarkers and aid in diagnosis of many chronic
diseases such as inflammatory bowel disease, diabetes and colorectal
cancer. Progress has been made towards developing ML models that predict
health outcomes from bacterial abundances, but reliable ML models are
scarce in part due to the flawed modeling methods. Furthermore, using
black box ML models has hindered the validation of microbial biomarkers.
To overcome these challenges, we benchmarked seven different ML models
that use fecal 16S rRNA sequences to predict the presence/absence of
colorectal cancer (CRC) lesions (n=490 patients, 261 controls and 229
cases). To show the effect of model selection, we assessed the accuracy,
interpretability, and computational efficiency of the following models:
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest, and extreme gradient boosting (XGBoost). The random
forest model was best at detecting CRC lesions with an AUROC of 0.695
but it was slow to train (101.3 h) and hard to interpret. Despite its
simplicity, L2-regularized logistic regression followed random forest in
predictive performance with an AUROC of 0.680, and it trained the
fastest (12 min). This study showed that we should choose ML models
based on our expectations of performance and interpretability as well as
our computational resources. It also established standards for modeling
pipelines of microbiome-associated ML models.

@@

\subsection{Background}\label{background}

Advances in sequencing technology and decreasing costs of generating 16S
rRNA gene sequences have allowed rapid exploration of human associated
microbiome and its health implications. Currently, the human microbiome
field is growing at an unprecedented rate and as a result, there is an
increasing demand for methods that identify associations between members
of the microbiome and human health. However, this is difficult as human
associated microbial communities are remarkably complex,
high-dimensional and uneven within and between individuals with the same
disease. It is unlikely that a single species can explain a disease.
Instead, subsets of those communities, in relation to one another and to
their host, account for differences in health outcomes.

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. Therefore,
researchers have started to explore the utility of ML models that use
microbiota associated biomarkers to predict human health and to
understand the microbial ecology of diseases such as liver cirrhosis,
colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
diabetes and others (1--11). However, currently the field's use of ML
lacks clarity and consistency on which methods are used and how these
methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably,
flawed ML practices are prevalent such as using ML pipelines where there
is no seperate held-out test dataset to evaluate model performance,
reporting few or only the best outcomes of different randomizations of
cross-validation and showing a disregard for large differences between
cross-validation and testing performances as well as large confidence
intervals of testing performances (4, 13--19). Moreover, there is a lack
of discussion on why a particular ML model is utilized. Recently, there
is a trend towards using more complex ML models such as random forest,
extreme gradient boosting and neural networks without a discussion on if
and how much model interpretibility is necessary for the study (11,
20--22).

The lack of transparency on modeling methodology and model selection
negatively impact model reproducibility and reliability. We need to
strive toward better machine learning practices by (1) implementing
consistent and reliable machine learning pipelines and (2) selecting ML
models that reflect the goal of the study as it will inform our
expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase a reliable ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and with a robust ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. CRC is the third most common form of cancer worldwide in which
the human-associated human microbiome is hypothesized to directly
contribute to its development. We built ML models using fecal 16S rRNA
gene sequences to predict patients with normal colons or patients with
colonic tumors which are called screen relevant neoplasias (SRN). The
study had 261 normal and 229 SRN samples. We established modeling
pipelines for L2-regularized logistic regression, L1 and L2 support
vector machines (SVM) with linear and radial basis function kernels, a
decision tree, random forest and XGBoost. Our ML pipeline utilized
held-out test data to evaluate generalization performance of each ML
model. The median test AUROC varied from 0.601 to 0.695. Random forest
had the highest median AUROC for detecting SRN. Despite its simplicity,
the L2-regularized logistic regression followed random forest in
performance. In terms of computational efficiency, L2 logistic
regression trained the fastest (0.202 hours, std ± 0.019), while XGBoost
took the longest (162.843 hours, std ± 3.986). We found that median
cross-validation and testing AUROC varied 0.021, which highlights the
importance of a seperate held-out test set and consistent preprocessing
of the data prior to evaluation. Aside from evaluating generalization
and classification performances for each of these models, this study
established standards for modeling pipelines of microbiome-associated
machine learning models.

@@

\textbf{Discriminative performance of the seven models.}

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using the area under the
receiver operating characteristic curve (AUROC) as the discriminative
performance metric {[}Figure 1{]}. Random forest had significantly
higher test AUROC values than the other models for detecting SRNs when
AUROC values were compared to the other six by Wilcoxon rank sum test (p
= 0.0011). The median AUROC of therandom forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another. They had median AUROC values
of 0.68 (IQR 0.055), 0.679 (IQR 0.053), 0.678 (IQR 0.056) and 0.668 (IQR
0.056) respectively. L1 SVM with linear kernel and decision tree had
significantly lower AUROC values than the other ML models with median
AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively {[}Figure
2{]}.

\textbf{Generalizability of the seven models.}

@@

\begin{itemize}
\item
  In this study we established a robust ML pipeline to use 16S rRNA
  sequence counts to predict a binary health outcome. We used a held-out
  test set to illustrate the difference between cross-validation and
  testing AUROC values. When the differences between cross-validation
  and test performance is low, this suggets the models will perform
  similar with similar data. In all seven models, the differences in
  cross-validation and testing AUROC values did not exceed 0.021 which
  suggests that these models will be able to test similar new data.
\item
  We also reported the results of 100 random dataset splits because
  there are differences in AUROC values of each of the seven models
  between different random data-splits. Our results showed that the
  testing AUROC values varied 0.23 on average, depending on the
  data-split. This result shows that you can get ``lucky'' in terms of
  prediction performance depending on the random split. Therefore, it is
  crucial to report all the findings over many random data-splits to get
  reliable performance metrics.
\item
  Our results also showed that we can choose different models for
  different reasons. Getting feature importances as weights and
  regression coefficients is the fastest and easiest interpratation.
  Similarly, linear models require less computational burden, fastest to
  train. Thinking about the applicability, and actionability of a model
  in clinical setting, we might want faster results.
\item
\item
  Suprisingly, l2 logistic regression does really well.
\item
  Computational efficiency and interp of logistic vs random forest
\item
  decision tree and l1 why so low?
\end{itemize}
